A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-31 text/abstract.Rmd
reduce abstract from 250 to 150 words (Cell requirement)

Cancer is a complex disease involving disrupted cellular metabolism, basic
biochemical processes, and the microenvironment. However, despite some generally
agreed upon unifying principles [@weinberg2000; @hallmarks_II], molecular
signatures remain largely indistinguishable from tissue-of-origin, presenting a
major barrier for precision health and individualized medicine. To address this
challenge, we reduce mutation data to disruptions in a select set of pathways
relevant to basic cell biology, from DNA replication to cellular communication.
Using dimensionality reduction techniques, we assign tumor samples into ten
clusters distinct from tissue-of-origin and largely free of bias from mutational
burden or clinical stage. We show that the clusters vary in prognosis by
modeling relative risk of death by cancer type and cluster. We identify
cluster-specific mutations in different tissues, demonstrating that
tissue-specific signatures contribute to common cellular phenotypes. Moreover,
germline risk genes involved in replication fidelity and genome stability are
equally distributed among clusters, contrary to the expectation that such genes
are avatars of molecular subtype. We investigate metastatic and non-metastatic
pathways, and show that most differences are cluster-specific. Some metastatic
pathways from one cluster are cluster-specific pathways from non-metastatic
tumors of another cluster, suggesting phenotypic convergence. Taken as a whole,
our observations suggest that common driver genes combine with tissue-specific
disruptions in tumor-promoting pathways to produce a limited number of distinct
molecular phenotypes. Thus, we present a coherent view of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

>>>>

Cancer affects cellular metabolism, biochemical processes, and the
microenvironment. However, despite some unifying principles, molecular
signatures remain largely indistinguishable from tissue-of-origin, presenting a
major barrier for precision health and individualized medicine. To address this,
we assign tumors into ten clusters based on pathways largely free of bias from
tissue-of-origin, mutational burden or stage. Germline risk genes are equally
distributed among clusters, contrary to the expectation that they determine
molecular subtype. We show that clusters vary in prognosis, modeling risk of
death by cancer type and cluster. We investigate metastasis, and show that most
differences are cluster-specific and complement other clusters, suggesting
convergence. Our observations suggest that driver genes combine with uncommon
genes in tumor-promoting pathways to produce distinct molecular phenotypes.
Thus, we present a coherent view of global tumor biology, and explain how common
cellular dysfunction might arise from tissue-specific mutations.
